Learning constraints in tabular data

### 2016-05-27 Paper/ECAI/tabular.tex
flashextract reworked

\section{Introduction}
Millions of people across the world use spreadsheets every day.
The tabular representation of the data is often intuitive, and the programming of functions in individual cells is quickly learned.
However, large and complex sheets (possibly with multiple tables and relations between them) can be hard to handle.
Many end-users lack the understanding of the underlying structures and dependencies in such sheets and the data they contain.
This is especially the case when spreadsheets have been exported from other software such as Enterprise Resource Planning (ERP) systems.
In this case, often a comma-separated values (CSV) format is used meaning that all formulas are lost, including inter-sheet formulas and relations.
%This limited understanding is especially the case with spreadsheets exported from other software such as ERP packages. In this case,
Even in manually created spreadsheets, it can be challenging to be consistent and correct with formulas across big spreadsheets.
For example the influential Reinhart-Rogoff economical paper ``Growth in a Time of Debt'' had some of its claims contested~\cite{flaw_excel}, after an investigation of the used Excel sheets was shown to contain some mistakes in formulae.

In this paper,  we investigate whether learning techniques can be used to infer constraints (formulas and other relations) from raw spreadsheet data.
This is a new and unconventional machine learning problem.
Consider the example in Figure~\ref{fig:main_example}, where the header's names already suggest the usage of spreadsheet operations such as \textit{average} or \textit{sum}.
Looking at the first row in Table~2 and the data in Table~1, it is clear that the values in this row were computed by summing the values in the corresponding column of Table~1.
Examining Table~1 also shows that computations are not only performed column-wise but row-wise as well: the cells in the column \textbf{Total} are obtained by taking the row-wise sum of the previous four columns.
This provides a flavor of how this problem is different from standard data mining settings, where the data is just in rows and variables are in columns. Here everything is mixed. The data is relational on the one hand, since we have multiple tables with relationships between them. And on the other hand, the data is most often mixed textual and numeric. %, since people tend to use spreadsheets for financial and accounting computations.
One can see that understanding and learning constraints in tabular data is hence a new and challenging problem for machine learning.

@@

\samuel{Move this up a paragraph}
The question that we answer in this paper is: is it possible to discover or reconstruct constraints (relations, functions) in flat tabular spreadsheet data?
To answer this question we contribute a general-purpose method and system, named \sname (from: Tabular Constraint Learner, pronounced ``tackle''), for discovering row-wise and column-wise constraints.
%Furthermore, the algorithm accepts declarative specifications of the constraints and predicates of interest (the "bias"). \tias{revise declarative claim?}
It operates directly on (headerless) tables of a spreadsheet in an unsupervised setting, as it reasons on that raw tabular data directly with no example constraint instantiations given. %, and discovers column- and row-wise constraints.
We demonstrate the utility of our approach in an experimental evaluation.
Moreover, we sketch additional application scenario's such as autocompletion and error detection.
\\\\
This paper is organized as follows.
Section \ref{sec:formalization} introduces concepts relevant to our approach. Section \ref{sec:approach} presents the problem statement and the approach. Section \ref{sec:evaluation} presents the evaluation of the approach. Section \ref{sec:applications} shows how our system can be used for applications. Section \ref{sec:related_work} presents an overview of the related work. Section \ref{sec:conclusions} provides conclusions.

@@

In line with this definition, we can use the following notation to refer to blocks: $B = T[\rangeto{a}{b},:]$ for a row-oriented block containing rows $a$ to $b$ in table $T$; and similarly $B = T[{:},\rangeto{a}{b}]$ for a column-oriented block.
We will refer to the \textit{vectors} of a block when we wish to refer to its rows/columns independently of their orientation.

@@

\item \textit{\CSignature} defines the requirements that the arguments of the predicate must satisfy.
This can concern any property of individual blocks as well as relations between properties of arguments, for example that the corresponding blocks must belong to the same table or have equal length.
In terms of logical and relational learning \cite{luc_book}, the \CSignature is known as the {\em bias} of the learner, it specifies when the arguments of a constraint are well-formed.

@@

It is helpful to see the analogy of constraint templates with first order logic (FOL) and constraint satisfaction.
\samuel{Is terms the right word? Shouldn't it be arguments, too?}
From a FOL perspective, the name of the constraint (e.g. \textit{SUM}) is just the predicate name, and the arguments $B_r$ and $B_x$ are the terms, which can be seen as either uninstantiated variables or as concrete values.
This also holds in our setting, where an instantiation of a variable corresponds to a concrete block. For example for $\ecrank{B_r}{B_x}$ and the spreadsheet in Figure~\ref{fig:main_example}, when we write $\ecrank{\range{T_1}{\rangeall}{8}}{\range{T_1}{\rangeall}{7}}$, then the value of $B_r$ is the $8$th vector in $T_1$: $B_r = \range{T_1}{\rangeall}{8} = [2,1,3,4]$ and the value of $B_x$ is the $7$th vector: $B_x = \range{T_1}{\rangeall}{7} = [1514, 1551, 691, 392]$.

@@

\subsubsection{Redundancies}
\label{sec:form:redundancies}
Depending on the application, some constraints in the solution to the tabular constraint learning problem may be considered \textit{redundant}. This is because constraints may be logically equivalent or may be implied by other constraints.

@@

%More formally, a constraint $c_1(G_1, ... G_n)$ \textit{implies} another constraint~$c_2(G'_1, ..., G'_m)$ whenever
%the  following statement holds: $$\forall G_i, G'_j : c_1(G_1, ... G_n) \rightarrow c_2(G'_1, ... G'_m).$$
%
%We shall only consider {\em range-restricted} implications, that is, we shall require that $\{G'_1, ..., G'_m\} \subseteq \{G_1, ..., G_n\}$.
%To simplify notation, and in line with logic programming, we shall often make the quantification implicit and simply write this as rules $ c_1(G_1, ... G_n) \rightarrow c_2(G'_1, ... G'_m)$.  We say that the first constraint is more specific than the second, and the second is more general.
%
%
%% where the second one ranges over a subsets of the arguments of the first, iff whenever the first constraint is true, the second must also be true. More formally, $\forall G'_1, ..., G'_n$ s.t. $\{G''_1, ..., G''_m\} \subseteq \{G'_1, ..., G'_n\},\, c_1(G'_1, ... G'_n) \rightarrow c_2(G''_1, ..., G''_m)$ll.
%
%\paragraph{Example}
%Consider the following implication:
%$$\ecperm{G} \rightarrow \ecalldiff{G}$$
%(it can be seen in Table~\ref{table:constraints} that both constraints only hold if $G$ is a vector).
%In this case the arguments are identical in both constraints.
%Given that the problem is to find the set of {\em all} constraints, and that the more general constraint is entailed
%by the more specific one, the $\ecdiff{G}$ constraint will be {\em redundant}.
%So, for most applications, it will be sufficient to output only the most specific one.
%
%A more involved example is
%$$\eclookup{G'_2}{G'_1}{H'_1}{H'_2} \rightarrow \ecfkey{G'_1}{H'_1}$$
%which states
%that a look-up constraint implies a foreign-key constraint. In the look-up constraint $\eclookup{G'_2}{G'_1}{H'_1}{H'_2}$ the values in the vector $G'_2$ are determined by looking up the corresponding value in vector $G'_1$ in vector $H'_1$ and returning the corresponding value in vector $H'_2$. In Figure~\ref{fig:main_example} we have $\eclookup{\range{T_{4}}{\rangeall}{3}}{\range{T_{4}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}}$. This implies a foreign-key constraint $\ecfkey{G'_1}{H'_1}$ where each of the values in vector $G'_1$ must appear in $H'_1$ for the look-up to work.

@@

We consider two cases of error detection.
The first setting is the \textit{online} detection of errors which is similar to the autocompletion setting.
However, we look at \textit{conflict} constraints, i.e. constraints that were learned on the snapshot data but do not agree with the newly added data~$\Delta D$.
Such conflict constraints can indicate errors in the input.
However, such conflicts can be wrongly caused by spurious constraints.
Therefore, conflicts should only be detected if the snapshot data is large enough.

@@

Third, our work borrows from the seminal work of Gulwani et al. \cite{flashfill} on program synthesis for tabular data that is incorporated in Microsoft’s Excel. In FlashFill, the end-user might work on a table containing the surnames (column A) and names (column B) of politicians. The first row might contain the values A1 = ``Blair'' and B1 = ``Tony'', and the user might then enter in cell C1 the value ``T.B.''. At that point FlashFill would generate a program that produces the output ``T.B.'' for the inputs ``Blair'' and ``Tony'', would also apply it to automatically generate values for the other politicians (rows) in the table. While it might need an extra example to deal with names such as ``Van Rompuy, Herman'' and determine whether it should output ``H.V.'' or ``H.V.R.'' \tias{only strings?}, FlashFill learns correct programs from very few examples. Flashfill has been extended to e.g. FlashExtract \cite{flashextract} to produce a set of normalized database tables \tias{can be more precise, does it add formulae? or this is real DB database, e.g a schema and PK/FK constraints?} starting from naively filled out spreadsheets. However, unlike our approach, Flashfill requires the user to identify explicitly the desired inputs/outputs from the function as well as to provide explicitly some positive and sometimes also negative examples.  In contrast, our approach is unsupervised, although it would be interesting to study whether it could be improved through user interaction and through user examples.  Another interesting open issue (for all of these techniques with the exception of logical and relational learning) is how to deal with possibly incomplete or noisy data.

>>>>

\section{Introduction}
Millions of people across the world use spreadsheets every day. The tabular representation of the data is often intuitive, and the programming of functions in individual cells is quickly learned. However, large and complex sheets (possibly with multiple tables and relations between different them) can be hard to handle. Many end-users lack the understanding of the underlying structures and dependencies in such sheets and the data they contain. Especially when spreadsheets have been exported from
other software such as Enterprise Resource Planning (ERP) systems; in this case, %, it is hard to understand the data as in this case,
often a comma-separated values (CSV) format is used meaning that all formula's are lost, including inter-sheet formulas and relations.
%This limited understanding is especially the case with spreadsheets exported from other software such as ERP packages. In this case,
Even in manually created spreadsheets, it can be challenging to be consistent and correct with formula's across big spreadsheets. For example the influential Reinhart-Rogoff economical paper ``Growth in a Time of Debt'' had some of its claims contested~\cite{flaw_excel}, after an investigation of the used Excel sheets was shown to contain some mistakes in formulae.

In this paper,  we investigate whether learning techniques can be used to infer constraints (formula's and other relations) from raw spreadsheet data.
This is a new and unconventional machine learning problem.
Consider the example in Figure~\ref{fig:main_example}, where the header's names already suggest the usage of spreadsheet operations such as \textit{average} or \textit{sum}.
Looking at the first row in Table~2 and the data in Table~1, it is clear that the values in this row were computed by summing the values in the corresponding column of Table~1.
Examining Table~1 also shows that computations are not only performed column-wise but row-wise as well: the cells in the column \textbf{Total} are obtained by taking the row-wise sum of the previous four columns.
This provides a flavor of how this problem is different from standard data mining settings, where the data is just in rows and variables are in columns. Here everything is mixed. The data is relational on the one hand, since we have multiple tables with relationships between them. And on the other hand, the data is most often mixed textual and numeric. %, since people tend to use spreadsheets for financial and accounting computations.
One can see that understanding and learning constraints in tabular data is hence a new and challenging problem for machine learning.

@@

%the field of Inductive Logic Programming has long... \tias{TODO for Sergey or Luc}. Additionally, the field of Constraint Programming has recently started investigating ways in which Constraint Satisfaction Problems can automatically be inferred from known solutions. Here, methods ranging from generate-and-test~\cite{todo:modelseeker}, concept learning~\cite{todo:conacq} and Inductive Logic Programming~\cite{todo:lallouet} have been used with encouraging results.

%Another line of work from which we draw inspiration is that of \tias{TODO for ?, flashfill et al}. We will consider the case where all data is given at once. However, as in the use case of flashfill, one could also use this technique to derive formula's interactively while a user is entering values in a cell \tias{TODO SAMUEL: proofread/refine last statement}.

The question that we answer in this paper is: is it possible to discover or reconstruct structural constraints (relations, functions) in flat tabular spreadsheet data?
To answer this question we contribute a general-purpose method and system, named \sname (pronounced as the word ``tackle''), for discovering row-wise and column-wise constraints.
%Furthermore, the algorithm accepts declarative specifications of the constraints and predicates of interest (the "bias"). \tias{revise declarative claim?}
It operates directly on (headerless) tables of a spreadsheet in an unsupervised setting, as it reasons on that raw tabular data directly with no example constraint instantiations given. %, and discovers column- and row-wise constraints.
We demonstrate the utility of our approach in an experimental evaluation.
Moreover, we sketch additional application scenario's such as autocompletion and error detection.
\\\\
This paper is organized as follows.
Section \ref{sec:formalization} introduces concepts relevant to our approach. Section \ref{sec:approach} presents the problem statement and the approach. Section \ref{sec:evaluation} presents the evaluation of the approach. Section \ref{sec:applications} shows how our system can be used for applications. Section \ref{sec:related_work} presents an overview of the related work. Section \ref{sec:conclusions} provides conclusions.

@@

In line with this definition, we can use the following notation to refer to blocks: $B = T[\rangeto{a}{b},:]$ for a row-oriented block containing rows $a$ to $b$ in table $T$; and similarly $B = T[{:},\rangeto{a}{b}]$ for a column-oriented block.
We will refer to the \textit{vectors} of a block when we wish to refer to its rows/columns independent of their orientation.

@@

\item the \textit{\CSignature} defines the requirements that the arguments of the predicate must satisfy.
This can concern any property of individual blocks as well as relations between properties of arguments, for example that the corresponding blocks must belong to the same table or have equal length.
In terms of logical and relational learning \cite{luc_book}, the \CSignature is known as the {\em bias} of the learner, it specifies when the constraint its arguments are well-formed.

@@

It is helpful to see the analogy of constraint templates with first order logic (FOL) and constraint satisfaction.
From a FOL perspective, the name of the constraint (e.g. \textit{SUM}) is just the predicate name, and the arguments $B_r$ and $B_x$ are the terms, which can be seen as either uninstantiated variables or as concrete values.
This also holds in our setting, where an instantiation of a variable corresponds to a concrete block. For example for $\ecrank{B_r}{B_x}$ and the spreadsheet in Figure~\ref{fig:main_example}, when we write $\ecrank{\range{T_1}{\rangeall}{8}}{\range{T_1}{\rangeall}{7}}$, then the value of $B_r$ is the $8$th vector in $T_1$: $B_r = \range{T_1}{\rangeall}{8} = [2,1,3,4]$ and the value of $B_x$ is the $7$th vector: $B_x = \range{T_1}{\rangeall}{7} = [1514, 1551, 691, 392]$.

@@

\paragraph{Redundancies}
Depending on the application, some constraints in the solution to the tabular constraint learning problem may be considered \textit{redundant}. This is because constraints may be logically equivalent or may be implied by other constraints.

@@

%
%More formally, a constraint $c_1(G_1, ... G_n)$ \textit{implies} another constraint~$c_2(G'_1, ..., G'_m)$ whenever
%the  following statement holds: $$\forall G_i, G'_j : c_1(G_1, ... G_n) \rightarrow c_2(G'_1, ... G'_m).$$
%
%We shall only consider {\em range-restricted} implications, that is, we shall require that $\{G'_1, ..., G'_m\} \subseteq \{G_1, ..., G_n\}$.
%To simplify notation, and in line with logic programming, we shall often make the quantification implicit and simply write this as rules $ c_1(G_1, ... G_n) \rightarrow c_2(G'_1, ... G'_m)$.  We say that the first constraint is more specific than the second, and the second is more general.
%
%
%% where the second one ranges over a subsets of the arguments of the first, iff whenever the first constraint is true, the second must also be true. More formally, $\forall G'_1, ..., G'_n$ s.t. $\{G''_1, ..., G''_m\} \subseteq \{G'_1, ..., G'_n\},\, c_1(G'_1, ... G'_n) \rightarrow c_2(G''_1, ..., G''_m)$ll.
%
%\paragraph{Example}
%Consider the following implication:
%$$\ecperm{G} \rightarrow \ecalldiff{G}$$
%(it can be seen in Table~\ref{table:constraints} that both constraints only hold if $G$ is a vector).
%In this case the arguments are identical in both constraints.
%Given that the problem is to find the set of {\em all} constraints, and that the more general constraint is entailed
%by the more specific one, the $\ecdiff{G}$ constraint will be {\em redundant}.
%So, for most applications, it will be sufficient to output only the most specific one.
%
%A more involved example is
%$$\eclookup{G'_2}{G'_1}{H'_1}{H'_2} \rightarrow \ecfkey{G'_1}{H'_1}$$
%which states
%that a look-up constraint implies a foreign-key constraint. In the look-up constraint $\eclookup{G'_2}{G'_1}{H'_1}{H'_2}$ the values in the vector $G'_2$ are determined by looking up the corresponding value in vector $G'_1$ in vector $H'_1$ and returning the corresponding value in vector $H'_2$. In Figure~\ref{fig:main_example} we have $\eclookup{\range{T_{4}}{\rangeall}{3}}{\range{T_{4}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}}$. This implies a foreign-key constraint $\ecfkey{G'_1}{H'_1}$ where each of the values in vector $G'_1$ must appear in $H'_1$ for the look-up to work.

@@

We consider two cases of error detection.
The first setting is the \textit{online} detection of errors which is similar to the autocompletion setting.
However, we look at \textit{conflict} constraints, i.e. constraints that were learned on the snapshot data but do not agree with the newly added data~$\Delta D$.
Such conflict constraints can indicate errors in the input.
However, such conflicts can be wrongly caused by spurious constraints.
Therefore, conflicts should only be detected if the snapshot data is large enough. An exact specification of what large enough means (i.e., how to measure the confidence in the constraint violation) is a separate application-dependent research question and will be investigated in the future work.

@@

Third, our work borrows from the seminal work of Gulwani et al. \cite{flashfill} on program synthesis for tabular data that is incorporated in Microsoft’s Excel. In FlashFill, the end-user might work on a table containing the surnames (column A) and names (column B) of politicians. The first row might contain the values A1 = ``Blair'' and B1 = ``Tony'', and the user might then enter in cell C1 the value ``T.B.''. At that point FlashFill would generate a program that produces the output ``T.B.'' for the inputs ``Blair'' and ``Tony'', would also apply it to automatically generate values for the other politicians (rows) in the table. While it might need an extra example to deal with names such as ``Van Rompuy, Herman'' and determine whether it should output ``H.V.'' or ``H.V.R.'' \tias{only strings?}, FlashFill learns correct programs from very few examples. Flashfill has been extended to e.g. FlashExtract \cite{flashextract} to produce a set of tables starting from a document (a txt, spreadsheet or HTML file) with a set of examples marked for extraction. However, unlike our approach, Flashfill requires the user to identify explicitly the desired inputs/outputs from the function as well as to provide explicitly some positive and sometimes also negative examples.  In contrast, our approach is unsupervised, although it would be interesting to study whether it could be improved through user interaction and through user examples.  Another interesting open issue (for all of these techniques with the exception of logical and relational learning) is how to deal with possibly incomplete or noisy data.
